Xolotl: An Intuitive and Approachable Neuron and Network Simulator for Research and Teaching

### 2018-08-17 paper/paper.tex
minor changes to text

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. This software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. The simulator's architecture allows for the interactive manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. xolotl is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code facilitate reproducible and auditable research. Its ease of use and rich visualization capabilites make it an attractive option in teaching environments. Finally, xolotl is written in a modular fashion, includes detailed tutorials and worked examples, and is freely available at https://github.com/sg-s/xolotl, enabling seamless integration into the workflow of other researchers.

@@

Xoltol is designed to streamline this process and allows for any parameter in any model to be manipulated using graphical sliders, with immediate, real-time feedback of its behavior. Any model in the simulator can be manipulated using

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_benchmark}
	\caption{Comparison of speed and accuracy of xolotl, NEURON and DynaSim. The top row shows simulations of a tonically-firing Hodgkin-Huxley model with three conductances with constant injected current (as in Fig. \ref{fig:figurehh}). The bottom row shows simulations of  a bursting stomatogastric ganglion neuron model with 8 conductances (as in Fig. \ref{fig:figurenetwork}). Ratio of run-time to simulation time (relative speed) as a function of simulation time step (A, B). Simulation error as a function of the step size (C, D).  Relative speed of integration as a function of the simulation length (E, F). Relative speed of integration, normalised by system size, as a function of the number of compartments simulated simultaneously (G, H). All benchmarks were performed on the same computer, and all simulators using fixed time-step integration methods. NEURON was run using the Python wrapper. }
	\label{fig:figurebenchmark}
\end{figure}

>>>>

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. This software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. The simulator's architecture allows for the interactive manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. xolotl is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code facilitate reproducible and auditable research. Its ease of use and rich visualization capabilities make it an attractive option in teaching environments. Finally, xolotl is written in a modular fashion, includes detailed tutorials and worked examples, and is freely available at https://github.com/sg-s/xolotl, enabling seamless integration into the workflow of other researchers.

@@

Xolotl is designed to streamline this process and allows for any parameter in any model to be manipulated using graphical sliders, with immediate, real-time feedback of its behavior. Any model in the simulator can be manipulated using

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_benchmark}
	\caption{Comparison of speed and accuracy of xolotl, NEURON and DynaSim. The top row shows simulations of a tonically-firing Hodgkin-Huxley model with three conductances with constant injected current (as in Fig. \ref{fig:figurehh}). The bottom row shows simulations of  a bursting stomatogastric ganglion neuron model with 8 conductances (as in Fig. \ref{fig:figurenetwork}). Ratio of run-time to simulation time (relative speed) as a function of simulation time step (A, B). Simulation error as a function of the step size (C, D).  Relative speed of integration as a function of the simulation length (E, F). Relative speed of integration, normalized by system size, as a function of the number of compartments simulated simultaneously (G, H). All benchmarks were performed on the same computer, and all simulators using fixed time-step integration methods. NEURON was run using the Python wrapper. }
	\label{fig:figurebenchmark}
\end{figure}
